Connecting inductive definitions, generic models and a modal multiverse for algebra and combinatorics

### 2022-09-18 slides-brixen2022b.tex
Slides finished

\setbeamertemplate{headline}{}

@@

  \bigskip
  %\includegraphics[height=0.32\textwidth]{olivia-lattices}
  \bigskip
  \bigskip
  \bigskip
  \bigskip

  \begin{tikzpicture}
    \def\R{8pt}
    \node (title) {\vbox{\vspace*{-0.0em}Connecting \hil{inductive definitions},
    \hil{generic models} and \\ a \hil{modal multiverse} for algebra and combinatorics
    \\[0.8em]}};
    \begin{pgfonlayer}{background}
      \draw[decorate, very thick, draw=mypurple]
        ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
        ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
        ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
        ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
        cycle;
    \end{pgfonlayer}
  \end{tikzpicture}

  \scriptsize
  \textit{-- an invitation --}
  \bigskip
  \bigskip
  \bigskip

  REDCOM: \\
  \emph{Reducing complexity in algebra, logic, combinatorics}
  \ \\
  September 19th, 2022
  \bigskip

@@

% \begin{document}

@@

@@

% \begin{document}

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
\begin{frame}{A topos-theoretic multiverse}
  \begin{block}{}
    \justifying
    \textbf{Def.} A statement~$\varphi$ holds \ldots
    \begin{enumerate}
      \item\small \hil{everywhere} ($\necessary\varphi$) iff
      it holds in every (Grothendieck) topos (over the current base topos).
      \item \hil{somewhere} ($\possible\varphi$) iff
      it holds in some positive topos.
      \item \hil{proximally} ($\xpossible\varphi$) iff
      it holds in some positive ouvert topos.
    \end{enumerate}
  \end{block}
  \pause

  \newcommand{\seqbad}{\only<3>{infinite descending sequence}\only<4->{bad set}}
  \begin{changemargin}{-2em}{-1.5em}
  \begin{itemize}
  \small
    \item \justifying
    \only<1-4>{A relation is almost-full\textsubscript{ind} \\
    iff every infinite sequence \emph{everywhere} is good \\
    iff the \emph{generic infinite sequence} is good \\
    iff the \emph{theory of infinite sequences} proves goodness.\medskip}
    \only<5->{A relation is almost-full\textsubscript{ind} iff every infinite
    sequence \emph{everywhere} is good.}
    \pause

    \item\justifying
    \only<3-4>{A relation is well-founded\textsubscript{ind} \\
    iff \emph{everywhere} there are no \seqbad s \\
    iff the \emph{generic \seqbad} validates~$\bot$ \\
    iff the \emph{theory} of \seqbad s proves~$\bot$.}
    \only<5->{A relation is well-founded\textsubscript{ind} iff
    \emph{nowhere} there are \seqbad s.}
    \pause
    \pause
    \pause

    \item \only<2-6>{A ring element~$x$ is nilpotent \\
    iff \emph{everywhere} it is contained in all prime ideals \\
    iff it is contained in the \emph{generic prime ideal} \\
    iff the \emph{theory} of prime ideals proves~$x \in \ppp$.}
    \only<7->{\mbox{A ring element is nilpotent iff all prime
    ideals \emph{everywhere} contain it.}}

@@

    \pause
    \item \emph{Somewhere,} the law of excluded middle holds.
  \end{itemize}
  \end{changemargin}
\end{frame}}

@@

% \begin{document}
\addtocounter{framenumber}{-1}
\begin{frame}{Extracting constructive content}
  \begin{block}{}
    \justifying
    \textbf{Prop.} [Dickson] If~$X$ and~$Y$ are almost-full\textsubscript{ind},
    so is~$X \times Y$.
  \end{block}
  \pause

  \vspace*{-0.3em}
  \emph{Proof.}\small
  \vspace*{-0.6em}
  \begin{enumerate}
  \item\justifying It suffices to verify that the \emph{generic infinite
  sequence}~$\gamma = (\alpha,\beta) : \NN \to X \times Y$ is good. \pause Since
  being good can be put as a geometric implication (in fact, a geometric
  formula) and since \textsc{lem} holds \emph{somewhere}, we may assume~\textsc{lem}.\pause
  \item By \textsc{lem} and well-foundedness, there is a minimal value~$\alpha(i_0)$
  among all values of~$\alpha$. \vspace*{-0.2em}\pause
  Similarly, there is a minimal value~$\alpha(i_1)$ among~$(\alpha(n))_{n >
  i_0}$, \pause a minimal value~$\alpha(i_2)$ among~$(\alpha(n))_{n > i_1}$, and so
  on. \pause By \emph{proximal dependent choice}, we can proximally collect these indices
  into a function~$i : \NN \to \NN$. \vspace*{-0.2em}\pause
  \item Switching~\textsc{lem} on again, there is a minimal value~$\beta(i(k_0))$
  among all values of~$\beta \circ i$. \pause Hence~$\gamma$ is good in view of
  \begin{align*}
    \alpha(i(k_0)) &\leq \alpha(i(k_0+1)), &
    \beta(i(k_0)) &\leq \beta(i(k_0+1)).
  \end{align*}
  \end{enumerate}
\end{frame}

@@

Attribution:
Blass(?)
Alexander
Henry

@@

explanation of generic models

>>>>

%\setbeamertemplate{headline}{}
\setbeamertemplate{headline}{%
  \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex]{}%
    \insertsectionnavigationhorizontal{\paperwidth}{}{}%
  \end{beamercolorbox}%
  \vskip-7pt%
}

@@

  \bigskip
  %\includegraphics[height=0.32\textwidth]{olivia-lattices}
  \bigskip
  \bigskip
  \bigskip

  \scriptsize
  \textit{-- an invitation --}

  \setbeamercolor{block body}{bg=black!100}
  \begin{block}{}
  \centering\normalsize\color{white}
  Connecting \hil{inductive definitions},
  \hil{generic models} and \\ a \hil{modal multiverse} for algebra and combinatorics
  \end{block}

%  \begin{tikzpicture}
%    \def\R{8pt}
%    \node (title) {\vbox{\vspace*{-0.0em}Connecting \hil{inductive definitions},
%    \hil{generic models} and \\ a \hil{modal multiverse} for algebra and combinatorics
%    \\[0.8em]}};
%    \begin{pgfonlayer}{background}
%      \draw[decorate, very thick, draw=mypurple]
%        ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
%        ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
%        ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
%        ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
%        cycle;
%    \end{pgfonlayer}
%  \end{tikzpicture}

  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip

  REDCOM: \\
  \emph{Reducing complexity in algebra, logic, combinatorics} \\
  \ \\
  Brixen \\
  September 19th, 2022
  \bigskip

@@

\section{Infinite data}

@@

\section{Constructive red flags}

@@

\section{The set-theoretic multiverse}

@@

\section{The topos-theoretic multiverse}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1-lighter}\end{minipage}}
\begin{frame}{Toposes and generic models}
  \vspace*{-0.2em}
  \begin{changemargin}{-1.2em}{-1.2em}
  \begin{enumerate}
    \item\justifying A \hil{(Grothendieck) topos} is a category of sheaves over some
    \hil{site}.

    {\small\emph{Examples.} $\Set$, $\Sh(X)$, $\Set[\TT]$.\par}
    \bigskip
    \pause

    \item The \hil{Kripke--Joyal semantics} defines what it means for a
    statement~$\varphi$ to hold ``internally in a topos~$\E$'', written~``$\E
    \models \varphi$''. This semantics is sound with respect to intuitionistic
    logic.

%    {\small
%    \begin{itemize}
%    \item\justifying A statement holds in~$\Set$ iff it is true.
%    \item A statement holds in the ``topos of double-negation sheaves'' iff its
%    double-negation translation is true.
%    \item A statement holds in~$\Sh(X)$ iff a globalized version of it holds for
%    continuous families on the
%    space~$X$.\end{itemize}\par}
    \bigskip
    \pause

%    \item Every topos supports \hil{mathematical reasoning}: \\
%    If~$\ \E \models \varphi\ $ and if~$\ \varphi$ entails~$\psi$
%    intuitionistically,\, then~$\E \models \psi$.

    \item Let~$\TT$ be a geometric theory. The \hil{classifying
    topos}~$\Set[\TT]$ contains the \hil{generic~$\TT$-model}~$U_\TT$. It is
    \hil{conservative} in that for geometric implications~$\varphi$, the
    following are equivalent:
    \begin{enumerate}
      \item The statement~$\varphi$ holds for~$U_\TT$ in~$\Set[\TT]$.
      \item The statement~$\varphi$ holds for every~$\TT$-model in every
      topos.
      \item The statement~$\varphi$ is provable modulo~$\TT$.
    \end{enumerate}

    \begin{center}
      \begin{tikzpicture}[ultra thick, node distance=7mm]
        \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40] (a) {$\ZZ$};
        \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40, right=of a] (b) {$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$};
        \node[regular polygon, regular polygon sides=5, draw=orange!80,
        fill=orange!20, right=of b, rounded corners=1pt, inner sep=0cm] (c) {$\O_X$};
        \node[star, rounded corners=1pt, star points=10, inner sep=0cm, draw=purple!80, fill=purple!20, right=of c] {$U_\TT$};
      \end{tikzpicture}
    \end{center}
  \end{enumerate}
  \end{changemargin}
\end{frame}}

\begin{frame}{Modal algebra and combinatorics}
  \begin{block}{}
    \justifying
    \textbf{Def.} A statement~$\varphi$ holds \ldots
    \begin{enumerate}
      \item\small \hil{everywhere} ($\necessary\varphi$) iff
      it holds in every (Grothendieck) topos (over the current base topos).
      \item \hil{somewhere} ($\possible\varphi$) iff
      it holds in some positive topos.
      \item \hil{proximally} ($\xpossible\varphi$) iff
      it holds in some positive ouvert topos.
    \end{enumerate}
  \end{block}
  \pause

  \newcommand{\seqbad}{\only<3>{infinite descending sequence}\only<4->{bad set}}
  \begin{changemargin}{-2em}{-1.5em}
  \begin{enumerate}
  \small
    \item \justifying
    \only<1-4>{A relation is almost-full\textsubscript{ind} \\
    iff every infinite sequence \emph{everywhere} is good \\
    iff the \emph{generic infinite sequence} is good \\
    iff the \emph{theory of infinite sequences} proves goodness.\medskip}
    \only<5->{A relation is almost-full\textsubscript{ind} iff every infinite
    sequence \emph{everywhere} is good.}
    \pause

    \item\justifying
    \only<3-4>{A relation is well-founded\textsubscript{ind} \\
    iff \emph{everywhere} there are no \seqbad s \\
    iff the \emph{generic \seqbad} validates~$\bot$ \\
    iff the \emph{theory of \seqbad s} proves~$\bot$.}
    \only<5->{A relation is well-founded\textsubscript{ind} iff
    \emph{nowhere} there are \seqbad s.}
    \pause
    \pause
    \pause

    \item \only<2-6>{A ring element~$x$ is nilpotent \\
    iff \emph{everywhere} it is contained in all prime ideals \\
    iff it is contained in the \emph{generic prime ideal} \\
    iff the \emph{theory of prime ideals} proves~$x \in \ppp$.}
    \only<7->{\mbox{A ring element is nilpotent iff all prime
    ideals \emph{everywhere} contain it.}}

@@

    \pause
    \item \emph{Somewhere,} the law of excluded middle holds.
  \end{enumerate}
  \end{changemargin}
\end{frame}

@@

\addtocounter{framenumber}{-1}
\begin{frame}{Extracting constructive content}
  \begin{block}{}
    \justifying
    \textbf{Prop.} [Dickson] If~$X$ and~$Y$ are almost-full\textsubscript{ind},
    so is~$X \times Y$.
  \end{block}
  \pause

  \vspace*{-0.3em}
  \emph{Proof.}\small
  \vspace*{-0.4em}
  \begin{enumerate}
    \item\justifying It suffices to verify that the \emph{generic infinite
    sequence}~$\gamma = (\alpha,\beta) : \NN \to X \times Y$ is good. \pause Since
    being good can be put as a geometric implication (in fact, a geometric
    formula) and since \textsc{lem} holds \emph{somewhere}, we may assume~\textsc{lem}.\pause

    \item By \textsc{lem} and well-foundedness, there is a minimal value~$\alpha(i_0)$
    among all values of~$\alpha$. \vspace*{-0.2em}\pause
    Similarly, there is a minimal value~$\alpha(i_1)$ among~$(\alpha(n))_{n >
    i_0}$, \pause a minimal value~$\alpha(i_2)$ among~$(\alpha(n))_{n > i_1}$, and so
    on. \pause By \emph{proximal dependent choice}, we can proximally collect these indices
    into a function~$i : \NN \to \NN$; this switches~\textsc{lem} off. \vspace*{-0.2em}\pause

    \item Switching~\textsc{lem} on again, there is a minimal value~$\beta(i(k_0))$
    among all values of~$\beta \circ i$. \pause Hence~$\gamma$ is good in view of
    \begin{align*}
      \alpha(i(k_0)) &\leq \alpha(i(k_0+1)), &
      \beta(i(k_0)) &\leq \beta(i(k_0+1)).
    \end{align*}
    \end{enumerate}
\end{frame}

@@

Attribution:
Blass(?)
Alexander
Henry
oE countable

@@

4: topos and generic models
* truth in Set, truth in double-negation sheaves, truth in Sh(X)
* generic ring, generic group, ...
